River networks dampen long-term hydrological signals of climate change

### 2017-04-11 drafts/PNAS/Chezik_PNAS_MS.tex
Organized
Removed Lingering Defunct Files and Cleaned Up For Reproducibility.

\usepackage{upgreek}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}s billions are spent adapting to climate change by such means as building dikes or improving stormwater drainage \cite{Narain:2011}, certain habitats and processes provide a natural defense system that mitigates climate change impacts \cite{Jones:2012}. For example, coastal habitats such as oyster reefs, mangrove forests, and eelgrass beds shield 67\% of the United States coastline and 1.4 million people from sea level rise and storms \cite{Arkema:2013}. Protection and enhancement of these natural systems has become a key component of many climate mitigation strategies \cite{Guerry:2015}, as they return multiple cost-effective services in contrast to engineered alternatives \cite{Jones:2012}. For instance, revitalization of the Mississippi delta not only improves storm protection, but also increases production and sustainability of fisheries and wildlife resources, protects the water supply, and reduces the impact of wastewater effluent \cite{Conservation:1998}. The value of climate buffering habitats stands only to increase as average global temperatures climb. The Paris Agreement of 2016 aims to limit global temperature rise to 1.5\degree{C} \cite{Hulme:2016}, a goal that concedes a minimum doubling of observed warming \cite{Hartmann:2013}; therefore, an urgent challenge and opportunity is discovering, maintaining and restoring systems that naturally buffer against the oncoming impacts of climate change.

In the face of changing global precipitation patterns \cite{Donat:2016}, river networks may offer an unappreciated defense against shifting flow regimes under climate change\cite{Hartmann:2013,Palmer:2009}. Earlier snowmelt \cite{Rauscher:2008}, reduced snow pack \cite{McCabe:2014}, shifts from snow to rain, and changes in the annual distribution of precipitation are all impacting stream flow \cite{Hartmann:2013}, resulting in flashier flows with increased frequency of flooding \cite{Hirabayashi:2013}, longer periods of drought and critically low flows for wildlife and agriculture \cite{Melillo:2014}. However, we hypothesize that large free-flowing river networks may naturally dampen these signals of climate change by integrating across varied landscapes and different manifestations of climate. Processes that aggregate across asynchronous components tend to dampen the aggregate, a process known as the portfolio effect \cite{Doak:1998}. Climate is expressed differently and asynchronously over the landscape \cite{George:2015} and on account of their branching architecture and directional flow, river networks integrate these varied climate manifestations \cite{Peterson:2013}, potentially dampening the local climatic response of sub-catchments. Large free-flowing rivers may have diverse climate portfolios, with sub-catchments acting as assets in their portfolio, thereby dampening local climate impacts. Portfolio effects in rivers are already known to dampen short-term variation \cite{Moore:2015,Yeakel:2014} but attenuation of long-term climate driven shifts have yet to be examined.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=17.8cm]{Fig1_Map.pdf}
	\caption{Climate trends in mean annual temperature (MAT) and mean annual precipitation (MAP) within the Fraser River basin overlayed on a digital elevation model of British Columbia, Canada. Flow gauge sites (dots) are scaled by the size of the contributing area. The map is projected in Albers BC UTMs to provide an equal area depiction of the region but labels are expressed in WGS84 latitude and longitude.}
\label{fig:1}
\end{figure*}

@@

For a river network to act as a portfolio of climate and dampen the effects of climate change, expressions of climate and their trends over time need to vary throughout the network. Using the Western North America Climate tool (ClimateWNA) \cite{Wang:2016}, we quantified the diversity in climate trends within the Fraser River basin over the years concurrent with our hydrology trend estimates (Fig.~\ref{fig:1}) (\textit{Materials and Methods}). The rate of warming varies across the watershed (from increases of 0.03 to 0.04\degree{C}$\cdot$year\textsuperscript{-1}).Further, there have been increases in precipitation towards the more northeastern portion.

This spatial variation in climate change means that catchments of different size integrate different amounts of climate variability. We summarized the climate variability integrated by each flow gauge by calculating the standard deviation in climate trends within their catchments. Temperature and precipitation are likely to drive changing hydrology, therefore we included variables in our climate index that capture changes in extreme and mean temperature, as well as changes in precipitation volume and physical state (snow vs. rain). The standard deviations of these climate variables were scaled between 0 and 1 and summed as a general climate index for each catchment. Following our hypothesis, this climate variability index increases with area (Fig.~\ref{fig:2}), asymptoting in catchments greater than 40,000~km\textsuperscript{2}.Thus, rivers that drain larger catchments integrated more climate variability. Larger areas contain greater landscape complexity and thus a greater variety of climatic expressions.

\begin{figure}[h]
\centering
\includegraphics[width=.8\linewidth]{Fig2_ClimPort.pdf}
	\caption{Climate variability index as a function of catchment area. The mean trend line (blue) was estimated using robust linear regression and standard error estimates (grey) were calculated using the weighted standard deviation estimate.}
\label{fig:2}
\end{figure}

@@

In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments (Fig.~\ref{fig:3}) \cite{Moore:2015}. This relationship between watershed size and trend certainty could pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework (\textit{Materials and Methods}) we simulated time-series for each site with no underlying trend. These simulations were generated by a random-walk process with the observed standard deviation and autocorrelation parameters ($\hat{\upsigma}$, $\hat{\upphi}$) at our sites. Using the same generalized least squares model applied to the observed data, we estimated trends for 1000 simulated time series at each site for each response variable (e.g., Fig.~S3). This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

\begin{figure*}[h]
\centering
\includegraphics[width=17.8cm]{Fig3_AnnFun.pdf}
	\caption{Annual flow trend attenuation within the Fraser River basin. (\textbf{Left}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large). Blue lines represent observed attenuation; orange and red represent simulated attenuation that is weaker and stronger than observed. (\textbf{Right}) Density plots show null model simulated variance exponents ($\hat{\updelta}$) and the proportion on either side of the observed variance exponent (blue). Flow metrics include long-term flow-timing shifts (change per decade in day-of-year (DOY) to half annual flow), where decreasing trends suggest more annual flow is occurring earlier in the year (a), and the percent change per decade in minimum (b), maximum (c) and median (d) annual flow. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:3}
\end{figure*}

@@

\subsection*{Seasonally Shifting Flow Trend Dampening} Regional flow-trends and attenuation strength varied by season. Analysis of monthly trends revealed that winter flows are getting higher and summer flows are getting lower over time, but that these trends have been more variable in smaller catchments (Fig.~\ref{fig:4}, \textit{SI Appendix}, Figs.~S1, S2). For example, winter maximum flows in small catchments ranged from nearly no change to dramatic 47\% increases per decade with decreases only as large as 9\%. Despite these extreme locations, on average the basin has only experienced moderate winter maximum flow increases of 5-9\%. Summer flows exhibited the opposite, with decreases in high flows over time. However, there was weak evidence of attenuation in the spring (Fig.~\ref{fig:4}, \textit{SI Appendix}, Figs.~S1,S2). Overall, climate portfolios tended to have greater attenuation during stable periods of the year and decreased during seasonal transitions (\textit{SI Appendix}, Fig.~S3). In the spring, the likelihood of network-driven attenuation of maximum flow trends was as low as 25\% (Fig.~\ref{fig:4}). The deterioration of the network's attenuation strength was coincident with the spring freshet, when snowmelt drives high flows across the basin. Synchronization events such as the freshet may subvert river network dampening mechanisms by homogenizing the region's response to seasonally driven climate shifts.

\begin{figure*}[h]
\centering
\includegraphics[width=17.8cm]{Fig4_MaxMonth.pdf}
	\caption{Monthly maximum flow trend attenuation within the Fraser River basin. (\textbf{Left}) Fraser River's basin-wide maximum-flow trend estimates (i.e., intercept = vertical grey lines) by month with density distributions of null-model simulations. Observed values falling further from the center of the density distribution suggest greater evidence for changes in maximum flow and a greater shift in magnitude. (\textbf{Center}) Observed monthly Fraser River maximum-flow variance exponent ($\hat{\updelta}$, blue) and associated density distribution of simulated $\hat{\updelta}$ estimates. Decimal values represent the percent of simulated data exhibiting weaker attenuation (yellow) than observed. (\textbf{Right}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large), for four seasonally representative months. These reflect months in the prior columns and describe the variation in percent change per decade of maximum flow among sites. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:4}
\end{figure*}

@@

\section*{Discussion}
Climate change is already causing economic and conservation challenges for river systems worldwide\cite{Palmer:2009,Pecl:2017}. For example, sockeye salmon in the Fraser River have seen as much as 80\% pre-spawn mortality in years with later run-off and elevated temperatures, leading managers to reduce or close the fishery (e.g., 2013 and 2015). In the absence of climate dampening the impacts of these climate extremes could have been dramatically worse. As climate change progresses, river hydrology will continue to shift, further stressing riverine ecosystems and subsequently demanding responsive management. However, river networks provide an underappreciated defense against these climate-change impacts as they are uniquely organized to leverage locally-filtered expressions of climate into stability. A simple product of network form and gravity, stability emerges as river networks integrate landscape heterogeneity, creating a downstream portfolio of climate that smooths local extremes and tempers long-term flow trends. Therefore, larger rivers should have flow regimes that are less sensitive to local climate trends.

@@

\matmethods{
\subsection*{Data} We downloaded daily flow data from Environment Canada's HYDAT database. Within HYDAT we selected Fraser River basin hydro-metric gauge stations that were not observed to be dam influenced and which collected data in each month between 1970 and 2007 where no month was missing more than 5 days data. We selected this 38-year range because this period maximized the number of gauge stations in the Fraser River basin (\textit{n}=55) operating concurrently over a long enough time period in which climate trends may be observable. We interpolated missing data using a linear model in log-space, but of the 13,879 days of data only 23 days were estimated. These daily data were smoothed using a five-day rolling average before summarizing into annual and monthly response variables. This smoothing technique was used to reduce the influence of erroneous and unusually extreme values. Because dams control flow and stabilize flow trends, we removed sites where daily flow exhibited unusually high autocorrelation and low variability relative to other sites of similar size, and were observed downstream of, and atypically close to, a dam (\textit{n}=3). The downstream distance of gauge stations from dams was determined using GIS spatial layers available in the \textit{BC Data Catalog}, maintained by the provincial government of British Columbia. Finally, we summarized the daily data into annual and monthly maximum, median and minimum flow estimates and the day-of-year to half-annual-flow.

\subsection*{River discharge trend analysis} We estimated annual and monthly flow trends for each response variable at each gauge station using a generalized least squares (GLS) model with an AR1 autocorrelation function:
\begin{equation}
  \mathrm{flow}_{s,t} = a_s + b_s \mathrm{year}_{s,t} + \epsilon_{s,t}, \quad 
  \epsilon_{s,t} \sim \mathcal{N}(\upphi \epsilon_{s,t-1}, \sigma_\mathrm{flow}^2) \label{eq1},
\end{equation}
where $\mathrm{flow}_{s,t}$ represents a flow metric [log(maximum), log(minimum), log(median), logit(day-of-year to half-annual-flow)] at each site ($s$) and time point ($t$),  and $\mathrm{year}_{s,t}$ represents the time in years (e.g., 1970, 1971, \ldots, 2007) with 1988 subtracted to approximately center the predictor. Parameters $b_{s}$ (slope) and $a_{s}$ (intercept) represent the estimated mean effect of time on $\mathrm{flow}_{s,t}$ and the estimate of $\mathrm{flow}_{s,t}$ at time zero (i.e., 1988) respectively. Error ($\epsilon_{s,t}$) of the current time step for a given site was allowed to be correlated with that of the previous time step by $\upphi$ and was assumed to be normally distributed with a variance of $\sigma^{2}$. To ensure our model estimates and simulations remained within the calendar year we logit transformed our flow-timing response variable (i.e., day-of-year to half-annual-flow) after scaling the data between 0 and 1 (i.e., dividing by 365).

\subsection*{Basin trend attenuation analysis} To measure network-driven dampening of climate signalling, we regressed site-specific trend estimates (e.g., \% change$\cdot$decade\textsuperscript{-1}) onto each site's catchment area using a generalized least squares model:
\begin{equation}
	\hat{b}_{s} = c + d\sqrt{\mathrm{area}_{s}} + \eta_{s}, \quad
  \eta_{s} \sim \mathcal{N}(0, f(\mathrm{area}_{s})) \label{eq2},
\end{equation}
where $\hat{b}_{s}$ represents a flow trend at a given site ($s$) and $\mathrm{area}_{s}$ represents site $s$'s watershed area. Fitted $d$ and $c$ parameters represent the mean effect of $\sqrt{\mathrm{watershed area}}$ on flow trends and the mean basin-wide flow trend at a theoretical watershed area of zero, respectively. To quantify attenuation, we modeled the change in trend variability with the change in catchment area using an exponential variance function \cite[p.~211]{Pinheiro:2000}: 
\begin{equation}
	f(\mathrm{area}_{s}) = \sigma_b^2 \exp(2\updelta\sqrt{\mathrm{area}_{s}}) \label{eq3},
\end{equation}
where the variance of the estimated error ($\eta$) changes exponentially with increasing $\sqrt{\mathrm{area}}$. This variance function allowed us to predict the range of flow trend values that would be expected as watershed area increased.

\subsection*{River Attenuation Null Model Simulations} In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments \cite{Moore:2015}. This relationship between watershed size and trend certainty may pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework we simulated time-series for each site with no underlying trend. These simulations were generated by a random walk process with the observed standard deviation and autocorrelation parameters ($\hat{\upsigma}$, $\hat{\upphi}$) at our sites. Using the same GLS model and AR1 correlation structure as applied to the observed data (eq. \ref{eq1}), we estimated trends for 1000 simulated time series at each site for each response variable (e.g., Fig.~S4). This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

We then applied equation \ref{eq2} to each of the 1000 basin-wide simulations (e.g. Fig.~S5), resulting in 1 observed variance exponent parameter ($\hat{\updelta}$) and 1000 simulated $\hat{\updelta}$ for each flow metric. By comparing our observed attenuation with our basin-wide null-model simulations we addressed the potential that more variable flows in smaller catchments contribute to the observed pattern of flow trends as a function of watershed area (Fig.~\ref{fig:3}). This null-model approach asks how likely our observation is due to a sampling effect versus a network portfolio effect.

\subsection*{Dampening summary statistics} We calculated attenuation certainty as the percentage of simulated $\hat{\updelta}$ that were less than the observed $\hat{\updelta}$. We calculated attenuation strength as the ratio between the standard deviation at the smallest and largest watersheds as defined by the exponential variance function (i.e., $\sqrt{f(\mathrm{area}_1)} / \sqrt{f(\mathrm{area}_2)}$). To compare the degree of attenuation observed with the null-simulated attenuation, we compared this ratio with the same ratio calculated from the null-simulated data (i.e., $\left( \sqrt{f(\mathrm{area}_1)} / \sqrt{f(\mathrm{area}_2)} \right) / \left( \sqrt{f(\mathrm{area}_{\mathrm{null} 1})} / \sqrt{f(\mathrm{area}_{\mathrm{null} 2})} \right)$). Finally, to estimate how the day-of-year to half-annual-flow was changing, we estimated the steepest point of each site's logistic curve and multiplied this slope value by 365 to revert from our 0--1 scale to our original annual range of 0--365. To get a decadal rate we multiplied these results by 3.8 (number of decades in 38 years). We estimated the steepest part of the logistic curve using the *divide by four* rule[@Gelman:2008] (dividing the coefficient by four equals the first derivitive of the logistic curve at its steepest point).

\subsection*{Climate} In order to develop a climate index for each flow-gauge catchment we needed to estimate climate trends across the Fraser River basin and then spatially summarise them by flow-gauge catchment. Using Whitebox Geospatial Analysis Tools, a freely available and open source geospatial analysis software \cite{Lindsay:2016}, we delineated the Fraser River basin and the catchment areas of each flow-gauge station. Digital elevation models provided by the provincial government of British Columbia, CA and state goverment of Washinton, USA were pre-processed with a breaching algorithm and stream burned to facilitate proper flow path and accumulation models. We then used Climate WNA to estimate historic climate values on an evenly spaced 1 km\textsuperscript{2} grid across the Fraser Basin for each year and month in our study. Climate variables included the mean annual temperature (MAT), and precipitation (MAP), extreme minimum (EMT) and maximum (EXT) temperature and precipitation as snow (PAS). Climate variable trends were calculated at each grid point using a modified eq.~\ref{eq1}, where time points were no longer considered and our response variable becomes $\mathrm{climate}_{s}$.

>>>>

\usepackage{upgreek}
\usepackage{gensymb}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}s billions are spent adapting to climate change by such means as building dikes or improving stormwater drainage \cite{Narain:2011}, certain habitats and processes provide a natural defense system that mitigates climate change impacts \cite{Jones:2012}. For example, coastal habitats such as oyster reefs, mangrove forests, and eelgrass beds shield 67\% of the United States coastline and 1.4 million people from sea level rise and storms \cite{Arkema:2013}. Protection and enhancement of these natural systems has become a key component of many climate mitigation strategies \cite{Guerry:2015}, as they return multiple cost-effective services in contrast to engineered alternatives \cite{Jones:2012}. For instance, revitalization of the Mississippi delta not only improves storm protection, but also increases production and sustainability of fisheries and wildlife resources, protects the water supply, and reduces the impact of wastewater effluent \cite{Conservation:1998}. The value of climate buffering habitats stands only to increase as average global temperatures climb. The Paris Agreement of 2016 aims to limit global temperature rise to 1.5\celsius~\cite{Hulme:2016}, a goal that concedes a minimum doubling of observed warming \cite{Hartmann:2013}; therefore, an urgent challenge and opportunity is discovering, maintaining and restoring systems that naturally buffer against the oncoming impacts of climate change.

In the face of changing global precipitation patterns \cite{Donat:2016}, river networks may offer an unappreciated defense against shifting flow regimes under climate change \cite{Hartmann:2013,Palmer:2009}. Earlier snowmelt \cite{Rauscher:2008}, reduced snow pack \cite{McCabe:2014}, shifts from snow to rain, and changes in the annual distribution of precipitation are all impacting stream flow \cite{Hartmann:2013}, resulting in flashier flows with increased frequency of flooding \cite{Hirabayashi:2013}, longer periods of drought and critically low flows for wildlife and agriculture \cite{Melillo:2014}. However, we hypothesize that large free-flowing river networks may naturally dampen these signals of climate change by integrating across varied landscapes and different manifestations of climate. Processes that aggregate across asynchronous components tend to dampen the aggregate, a process known as the portfolio effect \cite{Doak:1998}. Climate is expressed differently and asynchronously over the landscape \cite[e.g.][]{George:2015} and on account of their branching architecture and directional flow, river networks integrate these varied climate manifestations \cite{Peterson:2013}, potentially dampening the local climatic response of sub-catchments. Large free-flowing rivers may have diverse climate portfolios, with sub-catchments acting as assets in their portfolio, thereby dampening local climate impacts. Portfolio effects in rivers are already known to dampen short-term variation \cite{Moore:2015,Yeakel:2014} but attenuation of long-term climate driven shifts have yet to be examined.

\begin{figure*}[h]
\centering
\includegraphics[width=17.7cm]{Fig1_Map.pdf}
	\caption{Climate trends in mean annual temperature (MAT) and mean annual precipitation (MAP) within the Fraser River basin overlayed on a digital elevation model of British Columbia, Canada. Flow gauge sites (dots) are scaled by the size of the contributing area. The map is projected in Albers BC UTMs to provide an equal area depiction of the region but labels are expressed in WGS84 latitude and longitude.}
\label{fig:1}
\end{figure*}

@@

@@

For a river network to act as a portfolio of climate and dampen the effects of climate change, expressions of climate and their trends over time need to vary throughout the network. Using the Western North America Climate tool (ClimateWNA) \cite{Wang:2016}, we quantified the diversity in climate trends within the Fraser River basin over the years concurrent with our hydrology trend estimates (Fig.~\ref{fig:1}) (\textit{Materials and Methods}). The rate of warming varies across the watershed (from increases of 0.03 to 0.04\celsius$\cdot$year\textsuperscript{-1}). Furthermore, there is an observed decrease in preciptiation along the Canadian Rockies transitioning to increasingly wetter trends along the coast.

This spatial variation in climate change means that catchments of different size integrate different amounts of climate variability. We summarized the climate variability integrated by each flow gauge by calculating the standard deviation in climate trends within their catchments. Temperature and precipitation are likely to drive changing hydrology, therefore we included variables in our climate index that capture changes in extreme and mean temperature, as well as changes in precipitation volume and physical state (snow vs. rain). The standard deviations of these climate variables were scaled between 0 and 1 and summed as a general climate index for each catchment. Following our hypothesis, this climate variability index increases with area (Fig.~\ref{fig:2}), asymptoting in catchments greater than 40,000~km\textsuperscript{2}. Thus, rivers that drain larger catchments integrated more climate variability. Larger areas contain greater landscape complexity and thus a greater variety of climatic expressions.

\begin{figure}[h]
\centering
\includegraphics[width=8.69cm]{Fig2_ClimPort.pdf}
	\caption{Climate variability index as a function of catchment area. The mean trend line (blue) was estimated using robust linear regression and standard error estimates (SE) (2$\cdot$SE = grey) were calculated using the weighted standard deviation estimate.}
\label{fig:2}
\end{figure}

@@

In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments (Fig.~\ref{fig:3}) \cite{Moore:2015}. This relationship between watershed size and trend certainty could pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework (\textit{Materials and Methods}) we simulated time-series for each site with the observed standard deviation and autocorrelation parameters ($\hat{\sigma}$, $\hat{\upphi}$) but with no underlying trend. Using the same generalized least squares model applied to the observed data, we estimated trends for 1000 simulated time series at each site for each response variable. This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

\begin{figure*}[h]
\centering
\includegraphics[width=17.8cm, height=10.8cm]{Fig3_AnnFun.pdf}
	\caption{Annual flow trend attenuation within the Fraser River basin. (\textbf{Left}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large). Blue lines represent observed attenuation; orange and red represent simulated attenuation that is weaker and stronger than observed. (\textbf{Right}) Density plots show null model simulated variance exponents ($\hat{\updelta}$) and the proportion on either side of the observed variance exponent (blue). Flow metrics include long-term flow-timing shifts (change per decade in day-of-year (DOY) to half annual flow), where decreasing trends suggest more annual flow is occurring earlier in the year (a), and the percent change per decade in minimum (b), maximum (c) and median (d) annual flow. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:3}
\end{figure*}

@@

\subsection*{Seasonally Shifting Flow Trend Dampening} Regional flow-trends and attenuation strength varied by season. Analysis of monthly trends revealed that winter flows are getting higher and summer flows are getting lower over time, but that these trends have been more variable in smaller catchments (Fig.~\ref{fig:4}, \textit{SI Appendix}, Figs.~S1, S2). For example, winter maximum flows in small catchments ranged from nearly no change to dramatic 47\% increases per decade with decreases only as large as 9\%. Despite these extreme locations, on average the basin has only experienced moderate winter maximum flow increases of 5-9\%. Summer flows exhibited the opposite, with decreases in high flows over time. However, there was weak evidence of attenuation in the spring (Fig.~\ref{fig:4}, \textit{SI Appendix}, Figs.~S1,S2). Overall, climate portfolios tended to have greater attenuation during stable periods of the year and decreased during seasonal transitions. In the spring, the likelihood of network-driven attenuation of maximum flow trends was as low as 25\% (Fig.~\ref{fig:4}). The deterioration of the network's attenuation strength was coincident with the spring freshet, when snowmelt drives high flows across the basin. Synchronization events such as the freshet may subvert river network dampening mechanisms by homogenizing the region's response to seasonally driven climate shifts.

\begin{SCfigure*}[\sidecaptionrelwidth][h]
\centering
\includegraphics[width=11.4cm, height=8.5cm]{Fig4_MaxMonth.pdf}
	\caption{Monthly maximum flow trend attenuation within the Fraser River basin. (\textbf{Left}) Fraser River's basin-wide maximum-flow trend estimates (i.e., intercept = vertical grey lines) by month with density distributions of null-model simulations. Observed values falling further from the center of the density distribution suggest greater evidence for changes in maximum flow and a greater shift in magnitude. (\textbf{Center}) Observed monthly Fraser River maximum-flow variance exponent ($\hat{\updelta}$, blue) and associated density distribution of simulated $\hat{\updelta}$ estimates. Decimal values represent the percent of simulated data exhibiting weaker attenuation (yellow) than observed. (\textbf{Right}) Trend estimates $\pm$ one standard error (SE, grey) plotted against watershed area (km\textsuperscript{2}), colored by climate portfolio strength (green = small, blue = large), for four seasonally representative months. These reflect months in the prior columns and describe the variation in percent change per decade of maximum flow among sites. Simulated lines ignore variance in the intercept and slope to focus visually on attenuation.}
\label{fig:4}
\end{SCfigure*}

@@

\section*{Discussion}
Climate change is already causing economic and conservation challenges for river systems worldwide \cite{Palmer:2009,Pecl:2017}. For example, sockeye salmon in the Fraser River have seen as much as 80\% pre-spawn mortality in years with later run-off and elevated temperatures, leading managers to reduce or close the fishery (e.g., 2013 and 2015). In the absence of climate dampening, the impacts of these climate extremes could have been considerably worse. As climate change progresses, river hydrology will continue to shift, further stressing riverine ecosystems and subsequently demanding responsive management. However, river networks provide an underappreciated defense against these climate-change impacts as they are uniquely organized to leverage locally-filtered expressions of climate into stability. A simple product of network form and gravity, stability emerges as river networks integrate landscape heterogeneity, creating a downstream portfolio of climate that smooths local extremes and tempers long-term flow trends. Therefore, larger rivers should have flow regimes that are less sensitive to local climate trends.

@@

\matmethods{
\subsection*{Data} We downloaded daily flow data from Environment Canada's HYDAT database. Within HYDAT we selected Fraser River basin hydro-metric gauge stations that were not observed to be dam influenced and which collected data in each month between 1970 and 2007 where no month was missing more than 5 days data. We selected this 38-year range because this period maximized the number of gauge stations in the Fraser River basin (\textit{n}=55) operating concurrently over a long enough time period in which climate trends may be observable. We linearly interpolated missing data in log-space, but of the 13,879 days of data only 23 days were estimated. These daily data were smoothed using a five-day rolling average before summarizing into annual and monthly response variables. This smoothing technique was used to reduce the influence of erroneous and unusually extreme values. Because dams control flow and stabilize flow trends, we removed sites where daily flow exhibited unusually high autocorrelation and low variability relative to other sites of similar size, and sites that were observed downstream of, and atypically close to, a dam (\textit{n}=3). The downstream distance of gauge stations from dams was determined using GIS spatial layers available in the \textit{BC Data Catalog}, maintained by the provincial government of British Columbia. Finally, we summarized the daily data into annual and monthly maximum, median and minimum flow estimates and the day-of-year to half-annual-flow.

\subsection*{River discharge trend analysis} We estimated annual and monthly flow trends for each response variable at each gauge station using a generalized least squares (GLS) model with an AR1 autocorrelation function:
\begin{equation}
  \mathrm{flow}_{s,t} = a_s + b_s \mathrm{year}_{s,t} + \epsilon_{s,t}, \quad 
  \epsilon_{s,t} \sim \mathcal{N}(\upphi \epsilon_{s,t-1}, \sigma_\mathrm{flow}^2) \label{eq1},
\end{equation}
where $\mathrm{flow}_{s,t}$ represents a flow metric [log(maximum), log(minimum), log(median), logit(day-of-year to half-annual-flow)] at each site ($s$) and time point ($t$),  and $\mathrm{year}_{s,t}$ represents the time in years (e.g., 1970, 1971, \ldots, 2007) with 1988 subtracted to approximately center the predictor. Parameters $b_{s}$ (slope) and $a_{s}$ (intercept) represent the estimated mean effect of time on $\mathrm{flow}_{s,t}$ and the estimate of $\mathrm{flow}_{s,t}$ at time zero (i.e., 1988), respectively. Error ($\epsilon_{s,t}$) of the current time step for a given site was allowed to be correlated with that of the previous time step by $\upphi$ and was assumed to be normally distributed with a variance of $\sigma^{2}$. To ensure our model estimates and simulations remained within the calendar year we logit transformed our flow-timing response variable (i.e., day-of-year to half-annual-flow) after scaling the data between 0 and 1 (i.e., dividing by 365).

\subsection*{Basin trend attenuation analysis} To measure network-driven dampening of climate signalling, we regressed site-specific trend estimates (e.g., \% change$\cdot$decade\textsuperscript{-1}) onto each site's catchment area using a generalized least squares model:
\begin{equation}
	\hat{b}_{s} = c + d\sqrt{\mathrm{area}_{s}} + \eta_{s}, \quad
  \eta_{s} \sim \mathcal{N}(0, f(\mathrm{area}_{s})) \label{eq2},
\end{equation}
where $\hat{b}_{s}$ represents a flow trend at a given site ($s$) and $\mathrm{area}_{s}$ represents site $s$'s watershed area. Fitted $d$ and $c$ parameters represent the mean effect of watershed area ($\sqrt{\mathrm{area}}$), on flow trends and the mean basin-wide flow trend at a theoretical watershed area of zero, respectively. To quantify attenuation, we modeled the change in trend variability with the change in catchment area using an exponential variance function \cite[p.~211]{Pinheiro:2000}: 
\begin{equation}
	f(\mathrm{area}_{s}) = \sigma_b^2 \exp(2\updelta\sqrt{\mathrm{area}_{s}}) \label{eq3},
\end{equation}
where the variance of the estimated error ($\eta$) changes exponentially with increasing $\sqrt{\mathrm{area}}$. This variance function allowed us to predict the range of flow trend values that would be expected as watershed area increased.

\subsection*{River Attenuation Null Model Simulations} In the observed data, small watersheds exhibited greater variability around their trend estimates than large watersheds, likely because of greater short-term variation in small catchments \cite{Moore:2015}. This relationship between watershed size and trend certainty may pull small watershed trend estimates away from zero, thereby creating the appearance of decreasing trend variability among sites as watershed size increases. Using a null-hypothesis framework we simulated time-series with no underlying trend from each sites' fitted model, using the observed standard deviation and autocorrelation parameters ($\hat{\sigma}$, $\hat{\upphi}$). Using the same GLS model and AR1 correlation structure as applied to the observed data (eq. \ref{eq1}), we estimated trends for 1000 simulated time series at each site for each response variable (e.g., Fig.~S3). This simulation process, a form of parametric bootstrapping, created distributions of null-expectations with which we compared our observed results.

We then applied equation \ref{eq2} to each of the 1000 basin-wide simulations (e.g. Fig.~S4), resulting in 1 observed variance exponent parameter ($\hat{\updelta}$) and 1000 simulated $\hat{\updelta}$ for each flow metric. By comparing our observed attenuation with our basin-wide null-model simulations we addressed the potential that more variable flows in smaller catchments contribute to the observed pattern of flow trends as a function of watershed area (Fig.~\ref{fig:3}). This null-model approach asks how likely our observation is due to a sampling effect versus a network portfolio effect.

\subsection*{Dampening summary statistics} We calculated attenuation certainty as the percentage of simulated $\hat{\updelta}$ that were less than the observed $\hat{\updelta}$. We calculated attenuation strength as the ratio between the standard deviation at the smallest and largest watersheds as defined by the exponential variance function (i.e., $\mathrm{AS}=\sqrt{f(\mathrm{area}_{s=1})} / \sqrt{f(\mathrm{area}_{s=2})}$). To compare the degree of attenuation observed with the null-simulated attenuation, we compared this ratio with the same ratio calculated from the null-simulated data (i.e., $\mathrm{AS} / \mathrm{AS}_{NULL}$). Finally, to estimate how the day-of-year to half-annual-flow was changing, we estimated the steepest point of each site's logistic curve and multiplied this slope value by 365 to revert from our 0--1 scale to our original annual range of 0--365. To get a decadal rate we multiplied these results by 3.8 (number of decades in 38 years). We estimated the steepest part of the logistic curve using the \textit{divide by four} rule \cite{Gelman:2008} (dividing the coefficient by four equals the first derivitive of the logistic curve at its steepest point).

\subsection*{Climate} In order to develop a climate index for each flow-gauge catchment we needed to estimate climate trends across the Fraser River basin and then spatially summarise them by flow-gauge catchment. Using Whitebox Geospatial Analysis Tools, a freely available and open source geospatial analysis software \cite{Lindsay:2016}, we delineated the Fraser River basin and the catchment areas of each flow-gauge station. Digital elevation models provided by the provincial government of British Columbia, CA and state goverment of Washinton, USA were pre-processed with a breaching algorithm and stream burned to facilitate proper flow path and accumulation models. We then used Climate WNA to estimate historic climate values on an evenly spaced 1 km\textsuperscript{2} grid across the Fraser Basin for each year and month in our study. Climate variables included the mean annual temperature (MAT), and precipitation (MAP), extreme minimum (EMT) and maximum (EXT) temperature and precipitation as snow (PAS). Climate variable trends were calculated at each grid point using eq.~\ref{eq1} but where our response variable becomes $\mathrm{climate}_{s,t}$.

\subsection*{Code and Data Repository} All flow data analysis was done using R v3.3.2 (\url{www.r-project.org/}). Spatial analysis of watersheds was done using GDAL/OGR v1.11.5, and Whitebox GAT v3.4 \cite{Lindsay:2016} with visualization help from QGIS v2.18.2 (\url{www.qgis.org/}). Climate data were gathered using ClimateWNA v5.4 \cite{Wang:2016} and summary statitics were calculated using R and the \textit{rasterstats} module (\url{https://github.com/perrygeo/python-rasterstats.git}) in Python v2.7. All code can be found on \url{www.github.com/} at \url{http://github.com/kchezik/River-Network-Flow-Trends.git}. Raw HYDAT data can be obtained via Environment Canada (\url{http://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/}) or a simplified version can be found on GitHub.
}
